Contracts for unions and intersections are hard for Nickel

### 2021-05-28 paper.tex
Pass on section 5

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel]
let Port = contracts.fromPred (fun value =>
  builtins.isNum value && value % 1 == 0
  && value >= 0 && value <= 65353) in
let Configuration = {
  host | #Host,
  port | #Port,
} in
{host = "nixos.org", port = 80}
| #Configuration
\end{lstlisting}
  \end{center}
\caption{Contract composition}
\label{fig:contract-composition}
\end{figure}
\resolved{I wouldn't use nickel-lang.org unless you are
planning on putting something up there. Maybe nixos.org?}

@@

\subsection*{Flat contracts do not intersect}

@@

The papers that previously researched union and intersection
as dynamic checks have, in our eyes, different shortcomings
that makes them suboptimal solutions.

\subsection*{Keil and Thiemann}
\label{sec:keil-thiemann}
\unsure{Subsection title?}
Starting with the Keil and Thiemann work \cite{KeilThiemannUnionIntersection},
they noticed that, since union contracts should share information between
different application contexts, while intersection contracts should behave
independently between each application context, the system becomes much
simpler if every contract gets transformed into a union of intersections
of other contracts, by using the usual De Morgan laws,
and then eagerly opening up unions, but postponing intersection
contracts as much as possible.
\todo{This assumes the difficulty has been previously explained}

For instance, if the value \nickel{V} is wrapped with contract
\nickel{A & (B | C)}, then, before the contract is checked,
it will get opened up to something equivalent to
\nickel{(A & B) | (A & C)}.

This straightforward solutions makes composing the Keil and Thiemann
system with other existing systems more difficult; in the words
of Williams, Morris, and Wadler (see Section~\ref{sec:will-morr-wadl}):
``(\ldots) the monitoring semantics for contracts of intersection and union types given by Keil
and Thiemann are not uniform. (\ldots) If uniformity helps composition, then
special cases can hinder composition.''\cite{RootCauseOfBlame}
\info{(Yann) Does
it also make it less efficient? If $(A \& B)$ fails because of $B$, their
development causes to recheck the contract $A$ that has been duplicated?}
And it also could harm performance, or at least complicate an implementation,
since some sharing mechanism should be used to avoid checking the same contract,
over the same value, multiple times.\unsure{Arnaud says: this whole
  section is too abstract give concrete code and explain what goes
  wrong. It's fine to give code in Nickel syntax}

Another difficult issue Keil and Thiemann were able to solve, is the
fact that flat contracts and intersection do not usually play nice
together.
They provide two different sets of rules for their calculus,
the first one is non deterministic, whose purpose is to provide
semantics that are simple to understand.
The second one is a deterministic set of rules, that are intended
to be simple to implement.
This second set needs to perform operations
on the evaluation context, to decide whether or not to evaluate
a given contract.
\unsure{Worth showing how the non deterministic version works? (See section B.1.1)
try with an example}

We see two problems here, a minor one is the fact that the two
different semantics defined for the calculus are really far apart
one from the other, both on the strategies used for contract resolution,
as well as on the complexity levels.
\unsure{(Yann) Is it really a
    problem per se? It is common thing to do (a nice declarative system
    for proofs and an algorithmic one for implementation, proved equivalent).
    Maybe the point is that the algorithmic system is way more complex than the
declarative one
(Teo) reworded}.
A second, major one,
is the dependency added over the evaluation context, since, again,
it makes composing this, with other features desired on programming languages,
difficult, or even impossible
\unsure{(Yann) I think we need to substantiate this claim. If this
context dependency explains why CSE is invalid, maybe have a little example?
(Teo) I removed it, I don't think there's an example, mainly since
this context dependency is not part of the language. I do think it complicates
implementation of said feature}
\todo{(Yann) TODO: see if there's a low hanging complexity
    argument we can make (like it's quadratic in the number of open contracts or
whatever)}.

@@

\subsection*{Racket}
\label{sec:racket}

>>>>

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel]
let Port = contracts.fromPred (fun value =>
  builtins.isNum value && value % 1 == 0
  && value >= 0 && value <= 65353) in
let Configuration = {
  host | #Host,
  port | #Port,
} in
{host = "nixos.org", port = 80}
| #Configuration
\end{lstlisting}
  \end{center}
\caption{Contract composition}
\label{fig:contract-composition}
\end{figure}

@@

\subsection*{Flat contracts do not intersect}
\label{sec:flat-and-inter}

@@

The existing work that previously researched and implemented
union and intersection
as dynamic checks have, in our eyes, different shortcomings
that makes them suboptimal solutions.
In this section, we present informally present those issues,
starting with the foundational paper by Keil and Thiemann
\cite{KeilThiemannUnionIntersection}, then following to a recent
work by Williams, Morris, and Wadler \cite{RootCauseOfBlame}, and
finally going over some similar contracts available in the Racket
programming language.

\subsection*{Keil and Thiemann}
\label{sec:keil-thiemann}
\unsure{Subsection title?}
Keil and Thiemann noticed that, since union contracts should share information between
different application contexts, while intersection contracts should behave
independently between each application context, the system becomes much
simpler if every contract gets transformed into a union of intersections
of other contracts, by using the common De Morgan laws,
and then eagerly opening up unions, but postponing intersection
contracts as much as possible.
\todo{This assumes the difficulty has been previously explained}

For instance, if someone were to wrap a value \nickel{x}
in a contract as shown on the Original code snippet on Figure
\ref{fig:kt-contract-transf}, then, before actually executing that
contract, it would get transformed, and intersections would
be pushed inside the unions, as can be seen on the Transformed
code snippet on the same figure.

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Original]
x @ A & (B | C)
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Transformed]
x @ (A & B) | (A & C)
\end{lstlisting}
\caption{Contract transformation performed by the Keil and
Thiemann work}
\label{fig:kt-contract-transf}
\end{figure}

This straightforward solutions makes composing the Keil and Thiemann
solution with other existing systems more difficult; in the words
of Williams, Morris, and Wadler (see Section~\ref{sec:will-morr-wadl}):
``(\ldots) the monitoring semantics for contracts of intersection and union types given by Keil
and Thiemann are not uniform. (\ldots) If uniformity helps composition, then
special cases can hinder composition.''\cite{RootCauseOfBlame}
\info{(Yann) Does
it also make it less efficient? If $(A \& B)$ fails because of $B$, their
development causes to recheck the contract $A$ that has been duplicated?}
And it also could harm performance, or at least complicate an implementation,
since some sharing mechanism should be used to avoid checking the same contract,
over the same value, multiple times.\unsure{Arnaud says: this whole
  section is too abstract give concrete code and explain what goes
  wrong. It's fine to give code in Nickel syntax}

Another difficult issue Keil and Thiemann were able to solve, is the
fact that flat contracts and intersection do not usually play nice
together, as shown on Section \ref{sec:flat-and-inter}.
They provide two different sets of rules for their calculus,
the first one is non deterministic, and its purpose is to provide
semantics that are simple to understand.
With these rules, the interaction between flat contracts and
intersections becomes much simpler, since they are able to check
both sides of an intersection on separate threads.
This allows them to be sure that contracts from one branch of the intersection
won't get blamed while executing the other branch.

As an example, consider the programs shown on Figure
\ref{fig:kt-non-deterministic}, if the Original program
gets executed, it will spawn two instances of said program,
one with the first branch of the intersection (the First Program),
and the other one with the second one (the Second Program),
and these will get joined if they reach an equal execution point.

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Original]
x @ A & B
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=First Program]
x @ A
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Second Program]
x @ B
\end{lstlisting}
\caption{Non deterministic intersection contract opening}
\label{fig:kt-non-deterministic}
\end{figure}

The second set of rules is deterministic, and it is intended
to be simple to implement.
However, this second set needs to perform operations
on the evaluation context, to decide whether or not to evaluate
a given contract.
\footnote{You can check more on this topic on Section 5.1 of
\cite{KeilThiemannUnionIntersection}}

We see two problems here, a minor one is the fact that the two
different semantics defined for the calculus are really far apart
one from the other, both on the strategies used for contract resolution,
as well as on the complexity levels.
\resolved{(Yann) Is it really a
    problem per se? It is common thing to do (a nice declarative system
    for proofs and an algorithmic one for implementation, proved equivalent).
    Maybe the point is that the algorithmic system is way more complex than the
declarative one
(Teo) reworded}.
A second, major one,
is the dependency added over the evaluation context, since, again,
it makes composing this, with other features desired on programming languages,
difficult, or even impossible
\resolved{(Yann) I think we need to substantiate this claim. If this
context dependency explains why CSE is invalid, maybe have a little example?
(Teo) I removed it, I don't think there's an example, mainly since
this context dependency is not part of the language. I do think it complicates
implementation of said feature}
\todo{(Yann) TODO: see if there's a low hanging complexity
    argument we can make (like it's quadratic in the number of open contracts or
whatever)}.

@@

\subsection*{[WIP] Racket}
\label{sec:racket}
